### 2022-06-14 chapters/optimization/java-dir/java.tex
add : python

\section{Introduction}
As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners.%ROMAIN: Add source
Furthermore, if we take into consideration legacy applications, Java becomes the most used programming langues.%ROMAIN: Add source
In addition to its popularity, Java exhibits an interesting behaviour when it comes to energy consumption and performances, basically Java applications can be at the same time one of the most energy efficient or hungry solution.
As we have seen in the previous chapter, an inappropriate combination of parameters can drive Java applications from the top language to the bottom just by setting the wrong parameters.
Therefore, we wanted to dig deeper into this aspect of Java and study its runtime.
This chapter thus focuses on the impact of the runtime of Java applications on the energy consumption.

@@

\subsection{definition of the JVM}
\begin{figure}%[ht]
    \centering
    \includegraphics[width=\linewidth]{imgs/Jvm_architecture}
    \caption{JVM architecture}
    \label{fig:hotspot}
\end{figure}

\section{Experimental Protocol}\label{sec:meth}
To investigate the effect that could have the JVM distribution choice and/or parameters on software energy consumption, we conducted a wide set of experiments on a cluster of machines and using several established Java benchmarks and JVM configurations.

@@

\subsection{metrics and measurement}
Since the goal of this study is the green aspect Java virtual machines, our key metric will be the energy consumption of the job completed by each configuration.
In addition to the energy consumption, we collected additional metrics to explain the reasons behind the behaviour of each experiment.
Those extra metrics are:
\begin{itemize}
    \item execution time,
    \item number of threads.
\end{itemize}

@@

\begin{figure*}%[ht]
    \includegraphics[width=.9\linewidth]{imgs/powers_chetemi-2-dotty.pdf}
    \centering
    \captionsetup{justification=centering}
    \caption{Power consumption of \textsf{Dotty} as a service for \textsc{HotSpot}, \textsc{GraalVM} \& \textsc{J9}.}
    \label{fig:servicedotty}
\end{figure*}
\bigbreak
\begin{mdframed}[]
    To answer \textbf{RQ\,1}, we conclude that---while most of the JVM platforms perform similarly---we can cluster JVMs in 3 classes: \textsc{HotSpot}, \textsc{J9}, and \textsc{GraalVM}.
    The choice of one JVM of these classes can have a major impact on software energy consumption, that strongly depends on the application context.
    When it comes to the JVM version, latest releases tend to offer the lowest power consumption, but experimental features should be carefully configured, thus further questioning the impact of JVM parameters.
\end{mdframed}

>>>>

\section*{JAVA}
\section{Introduction}
As reported in the state of the art, Java is one of the most popular programming languages adopted by practitioners.%ROMAIN: Add source
Furthermore, if we take into consideration legacy applications, Java becomes the most used programming langues.%ROMAIN: Add source
In addition to its popularity, Java exhibits an interesting behaviour when it comes to energy consumption and performances, basically Java applications can be at the same time one of the most energy efficient or hungry solution.
As we have seen in the previous chapter, an inappropriate combination of parameters can drive Java applications from the top language to the bottom just by setting the wrong parameters.
Therefore, we wanted to dig deeper into this aspect of Java and study its runtime.
This chapter thus focuses on the impact of the runtime of Java applications on the energy consumption.

@@

\subsection{definition of the JVM}
% TODO : ADD definition here 
\begin{figure}%[ht]
    \centering
    \includegraphics[width=\linewidth]{imgs/Jvm_architecture}
    \caption{JVM architecture}
    \label{fig:hotspot}
\end{figure}

\section{Experimental Protocol}\label{sec:javameth}
To investigate the effect that could have the JVM distribution choice and/or parameters on software energy consumption, we conducted a wide set of experiments on a cluster of machines and using several established Java benchmarks and JVM configurations.

@@

\subsection{Metrics and measurement}
Since the goal of this study is the green aspect Java virtual machines, our key metric will be the energy consumption of the job completed by each configuration.
In addition to the energy consumption, we collected additional metrics to explain the reasons behind the behaviour of each experiment.
Those extra metrics are:
\begin{itemize}
    \item execution time,
    \item number of threads.
\end{itemize}

@@

\begin{figure*}%[ht]
    \includegraphics[width=.9\linewidth]{imgs/powers_chetemi-2-dotty.pdf}
    \centering
    \captionsetup{justification=centering}
    \caption{Power consumption of \textsf{Dotty} as a service for \textsc{HotSpot}, \textsc{GraalVM} \& \textsc{J9}.}
    \label{fig:servicedotty}
\end{figure*}
% \bigbreak
\begin{mdframed}[]
    To answer \textbf{RQ\,1}, we conclude that---while most of the JVM platforms perform similarly---we can cluster JVMs in 3 classes: \textsc{HotSpot}, \textsc{J9}, and \textsc{GraalVM}.
    The choice of one JVM of these classes can have a major impact on software energy consumption, that strongly depends on the application context.
    When it comes to the JVM version, latest releases tend to offer the lowest power consumption, but experimental features should be carefully configured, thus further questioning the impact of JVM parameters.
\end{mdframed}

### 2022-06-14 chapters/optimization/python-dir/python.tex
add : python

In particular, Noureddine~\emph{et~al.}~\cite{noureddine_preliminary_2012} in 2012, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
According to their experiments, an interpreted programming language like Python can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. Figure~\ref{fig:hannoi}).
% However, the operational conditions under which these measurements may threaten the validity of their results.

\begin{figure}[htbp]
    \includegraphics[width=\linewidth]{imgs/hannoiimplementation.png}
    \caption{Energy consumption of a recursive implementation of Tower of Hanoi program in different languages~\cite{noureddine_preliminary_2012}}
    \label{fig:hannoi}
\end{figure}
\subsection{python popularity}

However, Python seems to attract a large community of developers interested in data analysis, web development, system administration, machine learning---according to a survey conducted in 2018 by JetBrains---\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}}one can fear that the wide adoption of dynamic programming languages, like Python, in production may critically hamper the power consumption of ICT.

@@

In this chapter, we therefore explore the oblivious optimizations that can be applied to Python legacy applications in order to reduce their energy footprint.
As Python is widely adopted by software services deployed in public and private Cloud infrastructures, we believe that our contributions will benefit to wide diversity of legacy systems and not only favorably contribute to reduce the carbon emissions of ICT, but also reduce their cloud invoice for the resources consumed by these services.
More specifically, this chapter focuses on runtime optimizations that can adopted by developers to leverage the power consumption of Python applications.
We start by studuing the impact of some of programmers choice such as the type of data structures or loops, on the global energy consumption of the execution code.
then we discuss some other factors such as level of concurrency etc. later we we will talk about another non intrusive ways to optimize the energy consumption of this code. one kind of those optimizations includes alternative interpertersand some libraries that are dedicated to optimize the code without changing its structures such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.

@@

the simplicity, flexibility and the clarity of Python code led this programming language to a huge success. Therefore, a larger community and a wide range of support and libraries that allowed it to touch several fields like data analyses, machine learning, natural language processing, web development, astrophysics etc.
% python is not efficient 
% 1 Energy Efficiency Across Programming Languages: How Do Energy, Time, and Memory Relate?
%  \begin{figure}[hbt]
%     \includegraphics[width=\linewidth]{figures/hannoiimplementation.png}
%     \caption{Energy consumption of the recursive implementation of Tower of Hanoi program in different languages (using a base 10 logarithmic scale) \cite{noureddine_preliminary_2012}}      
%     \label{fig:hannoi}
% \end{figure}

However, and according to \cite{pinto_energy_2017,noureddine_preliminary_2012} and [1]. Python tends to be one more energy hungry . In figure \ref{fig:hannoi}, python consumes 30 times more than C or C++. The test was done on an implementation of \fnurl{hannoi test}{ https://en.wikipedia.org/wiki/Tower_of_Hanoi}  30 disks.

@@

In this chapter we will target the following research question :
\begin{enumerate}
    \item Do the data structres and controle structures impact the energy consumption ?
    \item how does the parallelism affect the energy behaviour of python programs ?
    \item are they any non intrusive methodes to reduce the imapct of the energy consumption of existing python softwares ?
    \item how can one apply those findings in the real world ?
\end{enumerate}

\subsection{use cases}
according to the survey made by Jetbrains company \footnote
{\fnurl{https://www.jetbrains.com/lp/python-developers-survey-2020}} data analysis and web develoepment are the top two purposes for python\ref{fig:usecase}

\begin{figure}[hbt]
    \centering
    \includegraphics[width=\linewidth]{imgs/python_use_cases}
    \caption{use cases of python }
    \label{fig:usecase}
\end{figure}

\subsubsection*{web developement}
first we wanted to test the energy consumption of python using on djagon,one of the most popular web frameworks.
therefore we create a simple website where we try to analyse the cost of a single request using different mechanisms to fetch the data from the database.
the idea behind this use case is to see wether the choice of the database and the orm ( object relational mapper) impacts the energy consumption and the perfromances of the website.

\subsubsection*{machine learning}
the other part is to study the impact of machine learning algorthims

%% ADD the work of alex 
\newpage
\section{experimetal protocole}

as we have discussed in the previous chapter \ref{chapter:benchmarking}. instead of running the tests the idea was to design a system that allows practiouners to reproduce and extends our tests. and then we use the same system to run answer some of researchs questions
%%%% probably ill resue the same things from other chapters

\section{python insights}

%% should i include the part with the exceptions ???
\subsection{conclusion}
This study has shown us, that the optimial way to reduce the energy consumption of the python code is to follow the guidelines and use the builtin functions, which is kinda happy news for the developers since they don't have to make extra effort to make their code green
%% except the parallellism part , which i dont know how to link it here yet

\newpage
\section{Python Runtimes}
One main inconvienent of the previous methode to reduce the energy consumption, is the hard work that should be done in order to alter the existing code base in order to redcue the energy consumption. To avoid such hustle we tried to find a non intrusive way to make the python code more ecofriendly without changing its structure. One aspect of python is the fact that it is an interptered language which led many works to create their own implementation of the interpeter to enhance one or many aspect of the python code. in the following section we will discuss the impact of those implementations on the energy consumption of python programs, and in which case, one should use a non conventional interpter to save the energy consumption of their program.

@@

@@

Despite the presence of the official docker images for the most of the runtimes, we prefered to build our own using the same reference image in order to remove any bias due to the Os used in the offical image.We used ArchLinux with the kernel version 4.9.184 as a base image.

\subsection{experimental setup}

\subsection{Power Measurements}

@@

\subsubsection{benchmarks}

\subsection{results and foundings}
As we can see in figure \ref{fig:tommi_all}, there is no evolution between cpython2 and cpython3
Intelpython and active python both folows the same behaviour. one can conclude that the work that have been on those interpreters is mainly
to improve a specific purpose, Active python claims that their version is focused on secuirty which explains the lack of some performances due to the introduction of more reonforcment
.Morver Intel published their version of python as a dedicated for machine learning. Unfortunately the tommti benchmark is a set that focuses on the general purpose programming
which does not reflect the performance of the machine learning.
Another aspect of a such behaviour might be due to the fact of the processors that were used in the testbed. and it might change in the future if we include the GPU part for the machine learning benchmarks
% reference the work of the intern 
As for nuitka. there were no optimization in the energy consumpotion despite the fact that is it a complier.
However if we dig through the nuitka mechanisms. they basically embed the python code with an interpreter.
Unllike nuitka shedskin exhebit a the best energy consumption pattern when it comes to the arithmetic operartions. One can conclude it is due to the fact of the native type of the variables. unlike the interpreters where they are treated as object in the begining.

the energy consumption.

\subsection{tools and package}

\subsection{Threads to validity}

\subsection{conclusion}

>>>>

In particular, Noureddine~\emph{et~al.}~\cite{noureddine_preliminary_2012} in 2012, and then Pereira~\emph{et~al.}~\cite{pereira_energy_2017} in 2017, conducted empirical power measurements on this topic, and both concluded that compiled programming languages overcome dynamic ones when it comes to power consumption.
According to their experiments, an interpreted programming language like Python can impose up to a $7,588\,\%$ energy overhead compared to C~\cite{pereira_energy_2017} (cf. Figure~\ref{fig:hannoi}).
In this chapter, we therefore explore the oblivious optimizations that can be applied to Python legacy applications in order to reduce their energy footprint.
As Python is widely adopted by software services deployed in public and private Cloud infrastructures, we believe that our contributions will benefit to wide diversity of legacy systems and not only favorably contribute to reduce the carbon emissions of ICT, but also reduce their cloud invoice for the resources consumed by these services.
More specifically, this chapter focuses on runtime optimizations that can adopted by developers to leverage the power consumption of Python applications.
We start by studuing the impact of some of programmers choice such as the type of data structures or loops, on the global energy consumption of the execution code.
then we discuss some other factors such as level of concurrency etc. later we we will talk about another non intrusive ways to optimize the energy consumption of this code. one kind of those optimizations includes alternative interpertersand some libraries that are dedicated to optimize the code without changing its structures such as \emph{ahead-of-time} (AOT) compilation and \emph{just-in-time} (JIT) libraries that are maintained by the community.
% However, the operational conditions under which these measurements may threaten the validity of their results.

\section{Motivation}
\subsection{python popularity}

Nowdays, Python seems to attract a large community of developers interested in data analysis, web development, system administration, machine learning---according to a survey conducted in 2018 by JetBrains---\footnote{\url{https://www.jetbrains.com/research/python-developers-survey-2018/}}one can fear that the wide adoption of dynamic programming languages, like Python, in production may critically hamper the power consumption of ICT.

@@

@@

the simplicity, flexibility and the clarity of Python code led this programming language to a huge success. Therefore, a larger community and a wide range of support and libraries that allowed it to touch several fields like data analyses, machine learning, natural language processing, web development, astrophysics etc.
\begin{figure}[htbp]
    \includegraphics[width=\linewidth]{imgs/hannoiimplementation.png}
    \caption{Energy consumption of a recursive implementation of Tower of Hanoi program in different languages~\cite{noureddine_preliminary_2012}}
    \label{fig:hannoi}
\end{figure}

However, and according to \cite{pinto_energy_2017} and \cite{noureddine_preliminary_2012}. Python tends to be one more energy hungry . In figure \ref{fig:hannoi}, python consumes 30 times more than C or C++. The test was done on an implementation of \fnurl{hannoi test}{ https://en.wikipedia.org/wiki/Tower_of_Hanoi}  30 disks.

@@

\subsection{use cases}
to reduce the energy consumption of python we started by targeting the main usage of this programming language, wich is revealed to be data sciences and web developpement. Figure~\ref{fig:usecase} illustrate a study made by the jetbrain company on python developers \footnote{\fnurl{https://www.jetbrains.com/lp/python-developers-survey-2020}}. In this survey, multiple multiple answers were accepted. 
\begin{figure}[hbt]
    \centering
    \includegraphics[width=\linewidth]{imgs/python_use_cases}
    \caption{use cases of python }
    \label{fig:usecase}
\end{figure}

\subsection*{Goal}
To reduce the energy consumption of python programs, we decided to start with studying its behaviour during the most common use cases. After this we will targtet the structures of the language itself in the hope to find some generic guidelines for more generic purpose as \citeauthor{hasan_energy_2016-1} did in their paper \cite{hasan_energy_2016-1}. After this we will measure the energy consumption of different python implementation for the goal of finding a non intrusive way to enhance this energy efficiency.

%% should we add Research questions ?

\newpage

\paragraph{Energy measurement}

%% should i include the part with the exceptions ??? naah no need so

%% except the parallellism part , which i dont know how to link it here yet 
\section{Workload} : 
for this chapter we are consediring 4 seperated, yet related studies . 
First we are going to see the energy behavriour of python within two study cases. Web servers and Machine learning. later we will dig deeper into the energy consumption of python basic structures where finally we will conclude with the impact of python interpterer on the energy consumption. 
Due to the lack of of support for most of  non conventional pyton interpeters, our main focus was only on micro benchmarks. 
Except pypy most of the python implementations don't support extra python libraries, Despite that most of the cases those extra implementations were made to optimize a specific library such as numba with numpy, or intelpython with machine learning algorthms.

\paragraph{preliminary studies}
for the first studies we used only the official version of python. because the goal was mainly to highlight the impact of the structure of code on the energy consumption.  
One main inconvienent of the previous methode to reduce the energy consumption, is the hard work that should be done in order to alter the existing code base in order to redcue the energy consumption. To avoid such hustle we tried to find a non intrusive way to make the python code more ecofriendly without changing its structure. One aspect of python is the fact that it is an interptered language which led many works to create their own implementation of the interpeter to enhance one or many aspect of the python code. in the following section we will discuss the impact of those implementations on the energy consumption of python programs, and in which case, one should use a non conventional interpter to save the energy consumption of their program.

@@

\section{experimetal protocole}

As we have discussed in the previous chapter \ref{chapter:benchmarking}. instead of running the tests, the idea was to design a system that allows practiouners to reproduce and extends our tests. and then we use the same system to run answer some of researchs question. 
%%%% probably ill resue the same things from other chapters 
\subsection{measurement context}
\paragraph{Hardware settings}

\paragraph{software settings}
For the sake of reproducibility, each experiment runs within a Docker container.
for each test we create a docker image. 
\subsection{Metrics}
Our focus will be mainly on CPU energy consumption because it is ten folds more than the DRAM one, since it is finit job benchmarking, time is highly correlated within the energy, and it will be only useful to explain certain energytical behaviour so we wont put a lot of focus on this metric.

%% basically i what i want to say is that we we care mostly about energy, because there are studies that were made on the performance, plus the DRAM is not that interested even within the heavy memory  benchmakrs 
% TODO : ADD DRAM graphs for vectors and matrix multiply

@@

Despite the presence of the official docker images for the most of the runtimes, we prefered to build our own using the same reference image in order to remove any bias due to the Os used in the offical image.We used ArchLinux with the kernel version 4.9.184 as a base image.
\section{results and finding}

% \subsubsection{benchmarks}

\subsubsection{web developement}
first we wanted to test the energy consumption of python using on djagon,one of the most popular web frameworks.
therefore we create a simple website where we try to analyse the cost of a single request using different mechanisms to fetch the data from the database.
the idea behind this use case is to see wether the choice of the database and the orm ( object relational mapper) impacts the energy consumption and the perfromances of the website.

\subsubsection{python and multiprocessing}
Figure \ref{fig:python_multiprocessing} shows the behaviour of python programs when we try to introduce the parallelism.
As we know python is a single threaded program thx to ghe GLS ( global lock system ) howerver due to the increase of the number of cores /threads per cpu it many libraries stared to take advanteage of this feature so most of them they will try to simulate the multithreading by using multiple instances pf the processor % TO VERIFY

\subsection{python insights}

\paragraph{conclusion}
This study has shown us, that the optimial way to reduce the energy consumption of the python code is to follow the guidelines and use the builtin functions, which is kinda happy news for the developers since they don't have to make extra effort to make their code green

\subsection{python runtimes}

As we can see in figure \ref{fig:tommi_all}, there is no evolution between cpython2 and cpython3
Intelpython and active python both folows the same behaviour. one can conclude that the work that have been on those interpreters is mainly
to improve a specific purpose, Active python claims that their version is focused on secuirty which explains the lack of some performances due to the introduction of more reonforcment
.Morver Intel published their version of python as a dedicated for machine learning. Unfortunately the tommti benchmark is a set that focuses on the general purpose programming
which does not reflect the performance of the machine learning.
Another aspect of a such behaviour might be due to the fact of the processors that were used in the testbed. and it might change in the future if we include the GPU part for the machine learning benchmarks
% reference the work of the intern 
As for nuitka. there were no optimization in the energy consumpotion despite the fact that is it a complier.
However if we dig through the nuitka mechanisms. they basically embed the python code with an interpreter.
Unllike nuitka shedskin exhebit a the best energy consumption pattern when it comes to the arithmetic operartions. One can conclude it is due to the fact of the native type of the variables. unlike the interpreters where they are treated as object in the begining.

@@

\section{Threads to validity}
\section{tools and contributions}

\section{conclusion}
